Supplementary Material

### 2024-03-19 article/paper_drought_Chile2023.qmd
update paper random forest landcover vs drought

title: Multi-temporal drought and its interaction with land cover change and vegetation productivity in continental Chile

@@

We wanted to explore the relationship between the trend in land cover classes and the trend in the drought indices. Regarding the trends of land cover obtained (see Section 3.4.5), we divide each trend (km2) per land cover class and macrozone by the total surface. That way, we relativize the trend regarding the total land cover surface type per zone. To the trends of the drought indices over Chile, we applied the mask of persistent land cover macroclass and aggregated the trend for land cover class and macrozone. We then explore visually the relationship between the trend of drought indices for the short- to long-term and the trend of land cover.

@@

We look at the relationship between drought index and land cover change in Figure 5 by comparing the trends in land cover change (in terms of the total surface area per land cover type and macrozone) and drought indexes. Figure 5 shows that the negative trends in cropland (-0.029) and savanna (-0.026) in “Norte Chico” are the highest and are associated with an increase trend in EDDI and a decreasing trend in SPI, SPEI, and SSI. On the contrary, the shrubland in “Centro” has an increase (0.04) linked to a decrease in SPI, SPEI, and SSI and to an increase in EDDI. In "Austral,” the positive trend in shrubland fits with the positive trend in all the drought indices. The rest of the land cover types in the macrozones show weak associations with water supply, water demand, or soil moisture drought indices.

```{r figTrendsLandDrought}
#| echo: false
#| fig-pos: "!ht"
#| fig-cap: "Relationship between the trend in land cover change and the trend in drought indices for the five macrozones. Vertical panels correspond to 1, 3, 6, 12, 24, and 36 months of the time scale by drought index. Horizontal panels show each drought index"
#| out-width: "\\textwidth"
knitr::include_graphics('../output/figs/points_landcover_drought_indices_trend_and_time_scale.png')
```

>>>>

title: The effects of multidimensional drought on land cover change and vegetation productivity in continental Chile

@@

We wanted to explore the relationship between the trend in land cover classes and the trend in the drought indices. For this purpose, in order to have more representative results, we conducted the analysis over sub-basins within continental Chile. We use 469 basins, which have a surface area between 0.0746 and  24,000 ($km^2$), and a median area of 1,249 ($km^2$). For each basin, we calculate the relative trend per land cover type, considering the proportion of the type relative to the total surface of the basin. Then, we extracted per basin the average trend of the drought indices SPI, SPEI, EDDI, SSI, and all their time scales 1, 3, 6, 12, 24, and 36. Also, we extracted the average trend in the proxy of vegetation productivity (zcNDVI). We wanted to analyze which drought indices and time scales have a major impact on changes in land cover type.

We have 25 predictors, which are drought indices and vegetation productivity. We analyzed the 25 predictors per type of landcover. For the analysis, we selected the method of random forest (@Ho1995).  Because it allows to find no linear relationship, it reduces overfitting and can derive the feature importance, which helps for a better understanding of the relationships. The importance of the variable is calculated by permuting out-of-bag (OOB) data per tree and calculating the mean standard error in the OOB. Then the same is done after permuting each predictor variable. Random forest uses multiple decision trees and allows for classification and regression.

We analyzed the 25 predictors per type of landcover, thus running six models. We used random forests for regression and trained 1000 forests. For more reliable results for the important variables, we resampled by creating ten folds, running a random forest per fold, and calculating the r-squared (rsq), root mean square error (RMSE), and variable importance ten times.

@@

According to Table \ref{tab-landcoverTrendRF}, barren land is the type of land cover that is most affected by the trend of drought indices that reach a rsq = 0.4. This is attributed to the fact that it is linked to short-term deficits in precipitation (SPI-6 and SPEI-6) and long-term AED (EDDI-12 and EDDI-24). The trend in drought indices accounts for about a quarter of the change in grassland, savanna, and shrubland types. These types are most affected by the increase in short-term water demand (EDDI-6). The changes in forests are strongly impacted by the increase in long-term water demand (EDDI-36). The trend in croplands cannot be attributed to variations in water demand, water supply, soil moisture, or vegetation productivity.

```{=latex}
\begin{table}[!ht]
\caption{The five most important trends of drought indices in estimating the landcover trend per land cover type and the r-squared (rsq) reached by each random forest model.}
\label{tab-landcoverTrendRF}
\includegraphics[]{../output/figs/table_importance_trends_landcover_vs_drought.png}
\end{table}
```

```{r fig-TrendsLandDrought}
#| echo: false
#| fig-pos: "!ht"
#| fig-cap: "Relationship between the trend in land cover change and the trend in drought indices for the five macrozones. Vertical panels correspond to 1, 3, 6, 12, 24, and 36 months of the time scale by drought index. Horizontal panels show each drought index"
#| out-width: "\\textwidth"
knitr::include_graphics('../output/figs/points_landcover_drought_indices_trend_and_time_scale.png')
```

We look at the relationship between drought index and land cover change in @fig-TrendsLandDrought by comparing the trends in land cover change (in terms of the total surface area per land cover type and macrozone) and drought indexes. @fig-TrendsLandDrought shows that the negative trends in cropland (-0.029) and savanna (-0.026) in “Norte Chico” are the highest and are associated with an increase trend in EDDI and a decreasing trend in SPI, SPEI, and SSI. On the contrary, the shrubland in “Centro” has an increase (0.04) linked to a decrease in SPI, SPEI, and SSI and to an increase in EDDI. In "Austral,” the positive trend in shrubland fits with the positive trend in all the drought indices. The rest of the land cover types in the macrozones show weak associations with water supply, water demand, or soil moisture drought indices.

### 2024-03-19 article/paper_drought_Chile2023.tex
update paper random forest landcover vs drought

\usepackage{lineno}\linenumbers \usepackage{multirow} \usepackage{lscape} \newcommand{\blandscape}{\begin{landscape}} \newcommand{\elandscape}{\end{landscape}}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Multi-temporal drought and its effects on land cover change and vegetation productivity in continental Chile},
  pdfauthor={Francisco Zambrano; Anton Vrieling; Francisco Fernández; Francisco Meza; Iongel Duran-Llacer; Alejandro Venegas-González; Nicolas Raab; Dylan Craven},
  pdfkeywords={drought, land cover change, satellite},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Multi-temporal drought and its effects on land cover change and
vegetation productivity in continental Chile}
\author[1]{Francisco Zambrano%
\corref{cor1}%
\fnref{fn1}}
 \ead{francisco.zambrano@umayor.cl} 
\author[2]{Anton Vrieling%
%
}
 \ead{a.vrieling@utwente.nl} 
\author[3]{Francisco Fernández%
%
}
 \ead{francisco.fernandez@uss.cl} 
\author[]{Francisco Meza%
%
}

@@

\author[]{Alejandro Venegas-González%
%
}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, enhanced, interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable, frame hidden]}{\end{tcolorbox}}\fi

@@

We wanted to explore the relationship between the trend in land cover
classes and the trend in the drought indices. Regarding the trends of
land cover obtained (see Section 3.4.5), we divide each trend (km2) per
land cover class and macrozone by the total surface. That way, we
relativize the trend regarding the total land cover surface type per
zone. To the trends of the drought indices over Chile, we applied the
mask of persistent land cover macroclass and aggregated the trend for
land cover class and macrozone. We then explore visually the
relationship between the trend of drought indices for the short- to
long-term and the trend of land cover.

@@

We look at the relationship between drought index and land cover change
in Figure 5 by comparing the trends in land cover change (in terms of
the total surface area per land cover type and macrozone) and drought
indexes. Figure 5 shows that the negative trends in cropland (-0.029)
and savanna (-0.026) in ``Norte Chico'' are the highest and are
associated with an increase trend in EDDI and a decreasing trend in SPI,
SPEI, and SSI. On the contrary, the shrubland in ``Centro'' has an
increase (0.04) linked to a decrease in SPI, SPEI, and SSI and to an
increase in EDDI. In ``Austral,'' the positive trend in shrubland fits
with the positive trend in all the drought indices. The rest of the land
cover types in the macrozones show weak associations with water supply,
water demand, or soil moisture drought indices.

@@

\caption{Relationship between the trend in land cover change and the
trend in drought indices for the five macrozones. Vertical panels
correspond to 1, 3, 6, 12, 24, and 36 months of the time scale by
drought index. Horizontal panels show each drought index}

@@

>>>>

\usepackage{lineno}\linenumbers \usepackage{multirow} \usepackage{lscape} \newcommand{\blandscape}{\begin{landscape}} \newcommand{\elandscape}{\end{landscape}}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The effects of multidimensional drought on land cover change and vegetation productivity in continental Chile},
  pdfauthor={Francisco Zambrano; Anton Vrieling; Francisco Meza; Iongel Duran-Llacer; Francisco Fernández; Alejandro Venegas-González; Nicolas Raab; Dylan Craven},
  pdfkeywords={drought, land cover change, satellite},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{The effects of multidimensional drought on land cover change and
vegetation productivity in continental Chile}
\author[1]{Francisco Zambrano%
\corref{cor1}%
\fnref{fn1}}
 \ead{francisco.zambrano@umayor.cl} 
\author[2]{Anton Vrieling%
%
}
 \ead{a.vrieling@utwente.nl} 
\author[]{Francisco Meza%
%
}

@@

\author[3]{Francisco Fernández%
%
}
 \ead{francisco.fernandez@uss.cl} 
\author[]{Alejandro Venegas-González%
%
}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, interior hidden, enhanced]}{\end{tcolorbox}}\fi

@@

We wanted to explore the relationship between the trend in land cover
classes and the trend in the drought indices. For this purpose, in order
to have more representative results, we conducted the analysis over
sub-basins within continental Chile. We use 469 basins, which have a
surface area between 0.0746 and 24,000 (\(km^2\)), and a median area of
1,249 (\(km^2\)). For each basin, we calculate the relative trend per
land cover type, considering the proportion of the type relative to the
total surface of the basin. Then, we extracted per basin the average
trend of the drought indices SPI, SPEI, EDDI, SSI, and all their time
scales 1, 3, 6, 12, 24, and 36. Also, we extracted the average trend in
the proxy of vegetation productivity (zcNDVI). We wanted to analyze
which drought indices and time scales have a major impact on changes in
land cover type.

We have 25 predictors, which are drought indices and vegetation
productivity. We analyzed the 25 predictors per type of landcover. For
the analysis, we selected the method of random forest (\citet{Ho1995}).
Because it allows to find no linear relationship, it reduces overfitting
and can derive the feature importance, which helps for a better
understanding of the relationships. The importance of the variable is
calculated by permuting out-of-bag (OOB) data per tree and calculating
the mean standard error in the OOB. Then the same is done after
permuting each predictor variable. Random forest uses multiple decision
trees and allows for classification and regression.

We analyzed the 25 predictors per type of landcover, thus running six
models. We used random forests for regression and trained 1000 forests.
For more reliable results for the important variables, we resampled by
creating ten folds, running a random forest per fold, and calculating
the r-squared (rsq), root mean square error (RMSE), and variable
importance ten times.

@@

According to Table \ref{tab-landcoverTrendRF}, barren land is the type
of land cover that is most affected by the trend of drought indices that
reach a rsq = 0.4. This is attributed to the fact that it is linked to
short-term deficits in precipitation (SPI-6 and SPEI-6) and long-term
AED (EDDI-12 and EDDI-24). The trend in drought indices accounts for
about a quarter of the change in grassland, savanna, and shrubland
types. These types are most affected by the increase in short-term water
demand (EDDI-6). The changes in forests are strongly impacted by the
increase in long-term water demand (EDDI-36). The trend in croplands
cannot be attributed to variations in water demand, water supply, soil
moisture, or vegetation productivity.

\begin{table}[!ht]
\caption{The five most important trends of drought indices in estimating the landcover trend per land cover type and the r-squared (rsq) reached by each random forest model.}
\label{tab-landcoverTrendRF}
\includegraphics[]{../output/figs/table_importance_trends_landcover_vs_drought.png}
\end{table}

@@

\caption{\label{fig-TrendsLandDrought}Relationship between the trend in
land cover change and the trend in drought indices for the five
macrozones. Vertical panels correspond to 1, 3, 6, 12, 24, and 36 months
of the time scale by drought index. Horizontal panels show each drought
index}

@@

We look at the relationship between drought index and land cover change
in Figure~\ref{fig-TrendsLandDrought} by comparing the trends in land
cover change (in terms of the total surface area per land cover type and
macrozone) and drought indexes. Figure~\ref{fig-TrendsLandDrought} shows
that the negative trends in cropland (-0.029) and savanna (-0.026) in
``Norte Chico'' are the highest and are associated with an increase
trend in EDDI and a decreasing trend in SPI, SPEI, and SSI. On the
contrary, the shrubland in ``Centro'' has an increase (0.04) linked to a
decrease in SPI, SPEI, and SSI and to an increase in EDDI. In
``Austral,'' the positive trend in shrubland fits with the positive
trend in all the drought indices. The rest of the land cover types in
the macrozones show weak associations with water supply, water demand,
or soil moisture drought indices.

### 2024-03-19 article/supplementary.qmd
update paper random forest landcover vs drought

---
title: "Supplementary Material"
format: pdf
editor: visual
---

@@

For vegetation, we obtained and use hereafter five macroclasses of land cover from IGBP MODIS: forest, shrubland, savanna, grassland, and croplands. Figure 1 c shows the spatial distribution of the macroclasses through Chile for the year 2022. The validation of IGBP2013 and IGBP2014 with LCChile reached near the same metrics of performance, having an accuracy of ~0.82 and a F1 score of ~0.66.

>>>>

---
title: "Supplementary Material"
author: "Francisco Zambrano"
format: 
  pdf:
    number-sections: true
editor: source
---

@@

For vegetation, we obtained and use hereafter five macroclasses of land cover from IGBP MODIS: forest, shrubland, savanna, grassland, and croplands. Figure 1 c shows the spatial distribution of the macroclasses through Chile for the year 2022. The validation of IGBP2013 and IGBP2014 with LCChile reached near the same metrics of performance, having an accuracy of \~0.82 and a F1 score of \~0.66.
